---
title: "Modelling censored data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{zero-inflation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


Censored data occurs where the observation falls either below or above some limit of detection. When the data falls below some limit of detection this is referred to as left-censoring, while when it falls above some limit of detection it is referred to as right-censoring. Interval censoring occurs when you know the upper and lower limit of the datapoint but not the actual value.

The `jsdmstan` package supports left-censoring in three families, the Gaussian (or normal) distribution , the lognormal distribution and the gamma distribution. Data can be simulated with left-censoring, which will return both the censored and uncensored response matrix. For fitting censored models, jsdmstan requires a matrix of the reponse plus an index matrix that says where in the matrix censoring occurs. For left-censoring the index matrix should be 1 when left-censored or 0 when not censored.


```{r setup}
library(jsdmstan)
```


## Example with simulated data

First, we can simulate data according to a lognormal distribution:

```{r}
set.seed(430150)
cens_dat <- jsdm_sim_data(N = 100, S = 15, K = 2, method = "mglmm", 
                        family = "lognormal", censor = "left", 
                        censor_points = sample(seq(0.05,1,0.05), 15, replace = TRUE))
```

We have to supply the censoring point per species to the data simulation. In this case we randomly choose those to be between 0.05 and 1. 

Now we can fit the model (lowering the number of iterations to improve runtime):

```{r}
cens_mod <- stan_jsdm(dat_list = cens_dat, family = "lognormal", method = "mglmm",
                      censoring = "left", iter = 2000, 
                      control = list(adapt_delta = 0.9))
cens_mod
```

No warnings are returned, or anything particularly concerning in terms of rhat or neff.

Now we can look at the data recovery, as there are some species that were simulated to have very extreme maximum values (check `max(nb_dat$Y)`), the overall distribution is hard to see using the default multi_pp_check options, so we use the statistics option and simultaneously check the recovery of the mean and the standard deviation:

```{r}
pp_check(cens_mod, plotfun="ecdf_overlay") + 
  ggplot2::coord_cartesian(xlim=c(0,5000))
multi_pp_check(cens_mod, plotfun = "ecdf_overlay")
```

We can see that the mean and standard deviation of the data are always nested within the model predictions, indicating a reasonable fit to the data.

We can also see if the model did well at recovering the parameters used within the data simulation:

```{r}
mcmc_plot(cens_mod, plotfun = "recover_hist", pars = "sigma",
          true = cens_dat$pars$sigma)
mcmc_plot(cens_mod, plotfun = "recover_hist", pars = paste0("betas[1,",1:15,"]"), #regexp = TRUE,
          true = cens_dat$pars$betas[1,])
mcmc_plot(cens_mod, plotfun = "recover_hist", pars = paste0("betas[2,",1:15,"]"), #regexp = TRUE,
          true = cens_dat$pars$betas[2,])
mcmc_plot(cens_mod, plotfun = "recover_hist", pars = paste0("betas[3,",1:15,"]"), #regexp = TRUE,
          true = cens_dat$pars$betas[3,])
mcmc_plot(cens_mod, plotfun = "recover_hist", pars = paste0("sigmas_species[",1:15,"]"), #regexp = TRUE,
          true = cens_dat$pars$sigmas_species)

```

